EGGs are adhesive!

### 2025-03-13 calco/main.tex
minori

\noindent
Thus, $L$ and $R$ are the lift- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet they are not affected by the rule itself.
%
If a category is adhesive, and the arrows of the rules are monomorphisms, then the presence of a match ensures that the
two pushouts exist, hence a rewriting step can be performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{xxx} extend the core framework, still ensuring that if the arrow 
of the rules are in $\mathcal{M}$,
%and the match is in $\mathcal{N}$, 
then the classical results from the theory of graph 
transformation can still be lifted~\cite{xxx}. 
If only the left-hand side belongs to $\mathcal{M}$, the current theory is stil under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being adhesive can be a daunting task. For this reason,  sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their adhesivity has not yet been settled.

@@

\noindent 
\parbox{9.5cm}{\begin{remark}\label{rem:obv}
Let $T_\Sigma$ be the functor sending $(\mathcal{H}, l)$ to $(T(\mathcal{H}), T(l))$.   Notice that, by definition, $(\mathcal{H}, l)$ is in $\EqTG_{\Sigma}$ if and only if $T(\mathcal{H})$ is in $\tg$. Thus there exists a functor $S_\Sigma$ as in the diagram on the right.
\end{remark}}\hfill \parbox{4cm}{\xymatrix@R=15pt{\EqTG_{\Sigma} \ar[r]_{J_\Sigma}  \ar[d]_{S_\Sigma}& \EqHyp_\Sigma \ar[d]^{T_\Sigma} \\ \tg \ar[r]^{I_\Sigma}& \hyp_{\Sigma}}}

@@

\section{Conclusions and further and related works}
\label{conclusioni}
The aim of our paper was to extend the theory of adhesive categories in order to include EGGS, 
a formalism for program optimisation 
%via a compact representation and 
%efficient implementation of equality saturation.
%
To do so, we revisited and generalised the notions of hyper-graph and term graph with equivalence, and
we extended it in order to capture EGGS as term graphs satisfying a suitable closure property.
Summing up, we proved that EGGs are objects of an $\mathcal{M}$-adhesive category $\GEqTGs$ 
and that optimisation steps are obtained via DPO rules (possibly with negative application conditions) whose 
left-hand side is in $\mathcal{M}$, and in fact an identity. 
%and via a match that is in $\mathcal{N}$. 
This fact allows for exploiting the properties 
of the adhesive framework when discussing about the optimisation process.

\emph{Future works.}
Our result on $\GEqTGs$ opens a few threads of research. The first is to check how the $\mathcal{M}$-adhesivity 
of EGGs can be pushed to model their rewriting via the double-pushout (DPO) approach. We have seen that 
the rules adopted in the literature of EGGs appears to be spans whose left-hand side is an identity and 
right-hand side is a regular mono (possibly with negative application 
conditions), and as such they fit the mould of rewriting on left-linear rewriting in $\mathcal{M}$-adhesive categories. 
%The consequences for usability on the restriction of the matches to arrows that are injective on equivalence classes
%though need to be further investigated.
%
Also, it still needs to be exploited how parallelism and causality, the key features for DPO rewriting
on adhesive categories, can be exploited in the context of EGGs implementation. 
Another venue for development is using the adhesive 
machinery to extend the EGGs formalism. In fact, most of the results
presented here for hypergraphs can be generalised to hierarchical hypergraphs, that is, 
hypergraphs with a hierarchy (a partial order) among 
edges~\cite{ghicaZan,CastelnovoGM24}, which is useful for adding structural information, such as encapsulation and sandboxing.

\emph{Related works.}
%\todo{espandere un poco}
Despite the interest they have ben raising as an efficient data structure, we are not aware of any attempt to provide an 
algebraic characterisation of EGGs, the only exception being~\cite{ghica}. We leave for future work an in-depth comparison 
among the two proposals, which appear to be related yet quite different.
%
In both cases, the key intuition is  to use string diagrams to represents (acyclic) term graphs. We adopted a
more down-to-Earth approach by equipping
concretely the nodes of a term graph with an equivalence relation, thus directly equivalent to the original presentation~\cite{DetlefsNS05},
at the same time extending and generalising~\cite{concur2006} in the contemporary jargon of adhesive categories.
The route chosen in~\cite{ghica} is to consider term graphs as arrows of a symmetric monoidal category,
and equipping them with an enrichment over semi-lattices, the resulting formalism being reminiscent of hierarchical hypergraphs.
Thus, the solution in~\cite{ghica}  is  more general than ours since it can be lifted to term graphs defined over categories other than
$\Set$.
%
At the same time, we both consider the DPO approach for rewriting, even if only our solution guarantees that the resulting category
is actually adhesive, thus allowing to exploit all the features of the framework.

>>>>

\noindent
Thus, $L$ and $R$ are the lift- and right-hand side of the rule, respectively, while $K$ witnesses those parts that must 
be present for the rule to be executed, yet they are not affected by the rule itself.
%
If a category is adhesive, and the arrows of the rules are monomorphisms, then the presence of a match ensures that the
two pushouts exist, hence a rewriting step can be performed.
%
The theory of $\mathcal{M}$-adhesivity~\cite{xxx} extend the core framework, still ensuring that if the arrows 
of the rules are in a suitable class $\mathcal{M}$ of monomorphisms
%and the match is in $\mathcal{N}$, 
then the classical results from the theory of graph 
transformation can still be obtained~\cite{xxx}. 
If only the left-hand side belongs to $\mathcal{M}$, the current theory is stil under development, as witnessed e.g.
by~\cite{BaldanC0G24}.
%
However, despite the elegance and effectiveness and the immediate advantages, 
%to be adhesive, 
proving that a given category satisfies the conditions 
for being adhesive can be a daunting task. For this reason,  sufficient criteria have been provided for the core 
framework, e.g. that every elementary topos is adhesive \cite{lack2006toposes}, as well as for the extended one of
$\mathcal{M}$-adhesivity~\cite{CastelnovoGM24}.
%
For some structures such as \emph{hypergraphs with equivalence} in~\cite{concur2006}, the question 
of their adhesivity has not yet been settled.

@@

\noindent 
\parbox{9.5cm}{\begin{remark}\label{rem:obv}
Let $T_\Sigma:  \EqHyp_{\Sigma} \to \hyp_{\Sigma}$ the forgetful functor lifting
$T:  \EqHyp \to \hyp$.
%
%$T_\Sigma = T\circ V_\Sigma$.   
Notice that, by definition, $(\mathcal{H}, l)$ is in $\EqTG_{\Sigma}$ amounts to say that $(T(\mathcal{H}), l)$ is in $\tg$. 
Thus there exists a functor $S_\Sigma$ as in the diagram on the right.
%Let $T_\Sigma$ be the functor sending $(\mathcal{H}, l)$ to $(T(\mathcal{H}), T(l))$.   Notice that, by definition, $(\mathcal{H}, l)$ is in $\EqTG_{\Sigma}$ if and only if $T(\mathcal{H})$ is in $\tg$. Thus there exists a functor $S_\Sigma$ as in the diagram on the right.
\end{remark}}
\hfill 
\parbox{4cm}{\xymatrix@R=15pt{\EqTG_{\Sigma} \ar[r]_{J_\Sigma}  \ar[d]_{S_\Sigma}& \EqHyp_\Sigma \ar[d]^{T_\Sigma} \\ \tg \ar[r]^{I_\Sigma}& \hyp_{\Sigma}}}

@@

\section{Conclusions and further and related works}
\label{conclusioni}
The aim of our paper was to extend the theory of $\mathcal{M}$-adhesive categories in order to include EGGS, 
a formalism for program optimisation 
%via a compact representation and 
%efficient implementation of equality saturation.
%
To do so, we revisited and generalised the notions of hyper-graph and term graph with equivalence, and
we extended it in order to capture EGGS as term graphs satisfying a suitable closure property.
Summing up, we proved that EGGs are objects of an $\mathcal{M}$-adhesive category $\GEqTGs$ 
and that optimisation steps are obtained via DPO rules (possibly with negative application conditions) whose 
left-hand side is in $\mathcal{M}$, and in fact an identity. 
%and via a match that is in $\mathcal{N}$. 
This fact allows for exploiting the properties 
of the $\mathcal{M}$-adhesive framework when discussing about the optimisation process.

\emph{Future works.}
Our result on $\GEqTGs$ opens a few threads of research. The first is to check how the $\mathcal{M}$-adhesivity 
of EGGs can be pushed to model their rewriting via the double-pushout (DPO) approach. We have seen that 
the rules adopted in the literature of EGGs appears to be spans whose left-hand side is an identity and 
right-hand side is a regular mono (possibly with negative application 
conditions), and as such they fit the mould of rewriting on left-linear rewriting in $\mathcal{M}$-adhesive categories. 
%The consequences for usability on the restriction of the matches to arrows that are injective on equivalence classes
%though need to be further investigated.
%
Also, it still needs to be exploited how parallelism and causality, the key features for DPO rewriting
on $\mathcal{M}$-adhesive categories, can be exploited in the context of EGGs implementation. 
Another venue for development is using the adhesive 
machinery to extend the EGGs formalism. In fact, most of the results
presented here for hypergraphs can be generalised to hierarchical hypergraphs, that is, 
hypergraphs with a hierarchy (a partial order) among 
edges~\cite{ghicaZan,CastelnovoGM24}, which is useful for adding structural information, such as encapsulation and sandboxing.

\emph{Related works.}
%\todo{espandere un poco}
Despite the interest they have ben raising as an efficient data structure, we are not aware of any attempt to provide an 
algebraic characterisation of EGGs, the only exception being~\cite{ghica}. We leave for future work an in-depth comparison 
among the two proposals, which appear to be related yet quite different.
%
In both cases, the key intuition is  to use string diagrams to represents (acyclic) term graphs. We adopted a
more down-to-Earth approach by equipping
concretely the nodes of a term graph with an equivalence relation, thus directly equivalent to the original presentation~\cite{DetlefsNS05},
at the same time extending and generalising~\cite{concur2006} in the contemporary jargon of adhesive categories.
The route chosen in~\cite{ghica} is to consider term graphs as arrows of a symmetric monoidal category,
and equipping them with an enrichment over semi-lattices, the resulting formalism being reminiscent of hierarchical hypergraphs.
Thus, the solution in~\cite{ghica}  is  more general than ours since it can be lifted to term graphs defined over categories other than
$\Set$.
%
At the same time, we both consider the DPO approach for rewriting, even if only our solution guarantees that the resulting category
is actually $\mathcal{M}$-adhesive, thus allowing to exploit all the features of the framework.
